Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto

### 2024-12-14 paper/paper.qmd
Final Paper Edits

---
title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
subtitle: "Population Density Drive Larger Budget Allocations meanwhile Average Household Income Drives Construction Projects"
author: 
  - Aviral Bhardwaj
thanks: "Code and data are available at: [https://github.com/Aviral-03/InfrastructureCausalModel-TO](https://github.com/Aviral-03/InfrastructureCausalModel-TO)." 
date: today
date-format: long
abstract: "This study investigates how population density and average household income affect budget allocations across Toronto’s 25 wards, and how these factors, along with budget allocation, influence construction projects. Using data from the city's \"Capital Budget Plan\", \"Ward Profiles\" and \"Active Building Permits\", we analyze spending patterns in key areas like related to infrastructre. Our analysis, guided by a causal model, presents a negative association between population density and construction activity, with some high-density wards receiving significantly higher budget allocations. Average household income and total budget allocations were positively correlated with the number of building permits. These findings highlight potential disparities in resource distribution, emphasizing the need for more equitable urban budget planning to address diverse community needs."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
prefer-html: true
---

@@

In urban governance, the equitable allocation of municipal budgets is essential to meet a city’s social, economic, and infrastructure needs. This challenge is especially pressing in Toronto, where 25 wards compete for limited resources amid a growing population. Understanding how demographic factors like population and income influence budget decisions remains underexplored. This study seeks to answer: How do population and income affect budget allocations across Toronto’s wards, and do these factors, along with budget allocation, impact the number of construction projects, a key indicator of infrastructure development?

Historical budget data shows uneven investment, particularly in high-density, lower-income areas, leading to disparities in infrastructure and services [@ferguson2024compromised]. Following her 2023 election, Mayor Olivia Chow prioritized housing, transit, and equity-focused initiatives [@toronto_2024_budget]. The 2024 budget reflects these priorities, with significant investments in affordable housing, transportation, and community safety, while emphasizing long-term infrastructure development through a 10-year capital plan. However, it remains unclear if these allocations are equitably distributed across wards or align with Toronto’s broader economic and demographic realities.

Major projects like the Ontario Line—a new subway connecting Toronto’s east and west ends—are expected to spur economic growth and development [@citeTransitExpansion]. Similarly, Olivia Chow’s "Sidewalks to Skylines" 10-year plan aims to create jobs and improve residents' quality of life [@citeSidewalksToSkylines]. As these initiatives unfold, equitable infrastructure investments will be vital for Toronto’s long-term prosperity. With more residents expected to spread across the city's wards, it is essential to evaluate whether current budget allocations align with the city’s broader development goals as the demographic characteristics of each ward changes.

This study uses data from the City of Toronto’s Open Data Portal, including demographic information from the "2021 Ward Profiles", "Capital Budget & Plan By Ward (10 yr Approved) 2021-2024", and "Building Applications and Permits". Our analysis, guided by a causal model, shows that both population density and average household income significantly influence budget allocations and the issuance of building permits, key indicators of infrastructure development. However, population density was identified as an exclusion restriction, as its inclusion in the Bayesian multiregression model revealed a counterintuitive negative relationship with the number of building permits. This suggests that unobserved factors may be influencing the link between population density and construction activity. After excluding population density from the model, we found that average household income and total budget allocations were positively correlated with building permits. This result aligns with expectations: wealthier wards with higher incomes receive larger budget allocations, leading to more construction projects.

The paper is organized as follows: [@sec-data] outlines the data sources and methodology used in the analysis. [@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, along with budget allocations, on construction projects. [@sec-results] presents key findings, including trends in budget allocations, the relationship between demographic factors and budget distribution, and the impact on construction projects. [@sec-discussion] explores the implications for urban governance and budget planning in Toronto. Finally, [@sec-limitations] addresses the study’s limitations and offers recommendations for future research. Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].

@@

Our research question and estimand analyse the causal relationships between population, income, budget allocation, and active construction projects ([@fig-causal-model]). Population and income serve as confounders, influencing both budget allocation and construction projects. Budget allocation acts as a mediator, connecting the confounders to the outcome, and as an instrumental variable, helping to estimate the effect of these predictors on the number of construction projects. The outcome variable is active construction projects, while population and income are the predictor variables. We aim to estimate the causal effect (estimand) of budget allocation on construction projects, while accounting for the influence of population and income. This paper highlights how budget decisions mediate the broader impact of socioeconomic factors on infrastructure development.

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-causal-model
#| fig-cap: "Causal model"
#| fig-subcap: "Population and income influence budget allocation and construction projects"
#| fig-pos: "h"

@@

The data, provided in CSV formats, was cleaned and analyzed using R [@citeR] programming language. The `readxl` [@readxl] package was used for reading Excel files. Other R packages used which includes `tidyverse` [@tidyverse], `styler` [@styler], and `dplyr` [@dplyr] for creating tables. The `ggplot2` [@ggplot2] and `kableExtra` [@kableExtra] were used for data visualization and table formatting. The `patchwork` [@patchwork] package was used for combining multiple plots, and `sf` [@sf] for spatial data analysis. The `lintr` [@citeLintr] package was used for code linting.

@@

These categories were selected because they represent significant areas of investment that directly impact the quality of life and well-being of Toronto residents. *State of Good Repair* focuses on maintaining and preserving existing infrastructure, *Growth Related* addresses the expansion and development of new infrastructure, and *Service Improvement and Enhancement* aims to enhance public services and amenities. By analyzing budget allocations in these categories, we can gain understand city's spending priorities and identify opportunities for more effective and equitable resource distribution. While also identifying potential disparities in funding across various wards of the city.

@@

The City Wards dataset [@toronto_city_wards], published by the City Clerk's Office and last updated on July 22, 2024, contains geographical information about each ward, including the ward ID, ward name, and ward boundary. These ward boundaries were decided as a part of `Bill 5, Better Local Government Act` in 2018, reducing the number of wards from 47 to 25 [@toronto_city_wards].

@@

The process of obtaining a building permit involves submitting an application to the City of Toronto, including necessary drawings, documents, and other forms based on the permit type. The Building Division reviews the application, and a Toronto Building Inspector ensures compliance with the Ontario Building Code, Zoning By-law, and other applicable regulations. Once approved, the permit is issued, allowing the applicant to commence construction or demolition.

@@

We selected this data-set to evaluate how budget allocations correlate with the number of building permits issued in each ward. The number of permits serves as a proxy for construction activity and development, highlighting the demand for infrastructure investment and capital expenditures. By examining the relationship between building permits and budget allocations, we can understand how construction activity influences resource distribution across the city's wards.

@@

There are several limitations to the analysis that need to be addressed. As discussed in [@sec-data], the datasets used in this study are publicly available from the City of Toronto. While they represents city's demographics, budget allocations, and building permits, they may not fully capture the complexity of Toronto’s infrastructure development and resource distribution. For example, the building permits dataset includes only active permits, which may not reflect the entire scope of construction activity. Additionally, the budget allocations dataset may exclude certain infrastructure projects and expenditures, leading to potential gaps in the analysis.

Furthermore, as detailed in [@sec-appendix], there are 45 unique statuses assigned to each submitted permit. This analysis focused on only one subset of these statuses, which may not fully capture the complexity of the building permit process and the various stages of construction. Future research could include additional permit statuses for a better understanding of construction activity. Another limitation is the lack of detailed data on specific projects within each budget category. This information could offer further understanding into why certain wards receive more or less funding.

Lastly, the model assumes that the relationships between population density, average household income, and budget allocations are linear, which may oversimplify these relationships. Additionally, the analysis assumes independence among wards, which may not be accurate due to shared resources and infrastructure projects. Future research could explore more advanced modeling techniques to address these non-linear relationships and dependencies.

>>>>

---
title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
subtitle: "Population density drive larger budget allocations meanwhile average household income drives construction projects"
author: 
  - Aviral Bhardwaj
thanks: "Code and data are available at: [https://github.com/Aviral-03/InfrastructureCausalModel-TO](https://github.com/Aviral-03/InfrastructureCausalModel-TO)." 
date: today
date-format: long
abstract: "This study examines the relationship between population density, average household income, and budget allocations across Toronto's 25 wards and investigates their combined influence on construction activity. Using data from the city's *Capital Budget Plan*, *Ward Profiles*, and *Active Building Permits*, we analyze spending patterns in key infrastructure-related areas. Employing a causal modeling approach, our findings reveal a negative association between population density and construction activity, despite some high-density wards receiving higher budget allocations. Additionally, average household income and total budget allocations exhibit a positive correlation with the number of building permits. These results underscore potential inequities in resource distribution, highlighting the need for data-driven and equitable urban budget planning to better address community needs."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
prefer-html: true
---

@@

In urban governance, the equitable allocation of municipal budgets is essential to meet a city’s social, economic, and infrastructure needs. This challenge is especially pressing in Toronto, where 25 wards compete for limited resources amid a growing population [@Ontario2024DemographicTrends]. Understanding how demographic factors influence budget decisions remains under-explored. Therefore, this study seeks to answer: How do demographic factors such as population and income affect budget allocations across Toronto’s wards, and can we assess their combined impact on construction activity, a key indicator of infrastructure development?

Looking back, Toronto's historical budget data have shown uneven investment, particularly in high-density, lower-income areas, leading to disparities in infrastructure and services [@ferguson2024compromised]. However, following 2023 Mayor election, Mayor Olivia Chow promise to prioritize housing, transit, and equity-focused initiatives [@toronto_2024_budget]. The 2024 budget prioritizes significant investments in affordable housing, transportation, and community safety, while also focusing on long-term infrastructure development through a 10-year capital plan. Notable projects, such as the *Ontario Line* subway expansion, are projected to drive economic growth and urban development [@citeTransitExpansion]. Similarly, the *Sidewalks to Skylines* 10-year plan aims to create jobs and enhance residents' quality of life [@citeSidewalksToSkylines]. As these initiatives progress, ensuring equitable infrastructure investments will be crucial for Toronto’s sustainable growth. With the city’s population increasingly distributed across its wards, it is imperative to assess whether current budget allocations effectively align with broader development objectives amidst shifting demographic trends.

This study analyzes data from the City of Toronto’s Open Data Portal, incorporating demographic information from the *2021 Ward Profiles*, budget details from the *Capital Budget & Plan By Ward (10-Year Approved) 2021-2024*, and construction data from *Building Applications and Permits*. Using a Bayesian multi-regression model guided by a causal framework, we found that both average household income and total budget allocations significantly influence the number of building permits. However, population density displayed an unexpected negative relationship with construction activity, suggesting the presence of unobserved factors affecting this link. To address this, we excluded population density as an explanatory variable and confirmed that higher-income wards receiving larger budget allocations corresponded to more construction activity. These findings highlight the importance of understanding how socioeconomic and budgetary factors shape urban development, emphasizing the need for equitable resource distribution to support balanced growth across Toronto’s wards.

The paper is organized as follows: [@sec-data] outlines the data sources and methodology used in the analysis. [@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, following the causal model ([@fig-causal-model]). [@sec-results] presents key findings, including trends in budget allocations, the relationship between demographic factors and budget distribution, and the impact on construction projects. [@sec-discussion] explores the implications for urban governance and budget planning in Toronto. Finally, [@sec-limitations] addresses the study’s limitations and offers recommendations for future research. Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].

@@

Our research investigates the causal relationships between population, income, budget allocation, and active construction projects, as illustrated in [@fig-causal-model]. Population and income are key confounders, influencing both budget allocation and the number of construction projects. Budget allocation serves a dual role: as a mediator connecting population and income to construction projects and as an instrumental variable, helping to estimate their effects on construction activity. The outcome variable is the number of active construction projects, while population and income are the primary predictors. Our goal is to estimate the causal effect of budget allocation on construction activity, accounting for the confounding effects of population and income.

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-causal-model
#| fig-cap: "Causal model: Population and income influence budget allocation and construction projects"
#| fig-pos: "h"
#| fig.align: "center"

@@

The data, provided in CSV formats, was cleaned and analyzed using R [@citeR] programming language. The `readxl` [@readxl] package was used for reading Excel files. Other R packages used includes `tidyverse` [@tidyverse], `styler` [@styler], and `dplyr` [@dplyr] for creating tables. `ggplot2` [@ggplot2] and `kableExtra` [@kableExtra] were used for data visualization and table formatting. The `patchwork` [@patchwork] package was used for combining multiple plots, and `sf` [@sf] for spatial data analysis. For model estimation, the `rstanarm` [@citerstanarm] package was used for Bayesian modeling, and `bayesplot` [@citeBayesPlot] for visualizing the results. The `lintr` [@citeLintr] package was used for code linting.

@@

Since our analysis aims to understand how budget allocations influence construction activity and infrastructure development, we selected these three categories as they are most relevant to capital expenditures and infrastructure projects. *State of Good Repair* focuses on maintaining and preserving existing infrastructure, *Growth Related* addresses the expansion and development of new infrastructure, and *Service Improvement and Enhancement* aims to enhance public services and amenities.

@@

The City Wards dataset [@toronto_city_wards], published by the City Clerk's Office and last updated on July 22, 2024, contains geographical information about each ward, including the ward ID, ward name, and ward boundary. These ward boundaries were decided as a part of *Bill 5, Better Local Government Act* in 2018, reducing the number of wards from 47 to 25 [@toronto_city_wards].

@@

The process of obtaining a building permit involves submitting an application to the City of Toronto, including necessary drawings, documents, and other forms based on the permit type. The Building Division reviews the application, and a Toronto Building Inspector ensures compliance with the Ontario Building Code, Zoning By-law, and other applicable regulations. Once approved, the permit is issued, allowing the applicant to commence construction or demolition, and thus data is saved in the system as an active permit.

@@

We selected this data-set to evaluate how our confounders, population and income, influence construction activity, and mediator, budget allocation, correlate with the number of building permits issued in each ward. The number of permits serves as a proxy for construction activity and development, highlighting the demand for infrastructure investment and capital expenditures. By examining the relationship between building permits and outlined variables, we can understand how construction activity influences resource distribution across the city's wards.

@@

This analysis has several limitations that warrant consideration. As discussed in [@sec-data], the datasets used, sourced from the City of Toronto, provide insights into the city’s demographics, budget allocations, and building permits. However, they may not fully capture the complexity of Toronto's infrastructure development and resource distribution.

For instance, in [@sec-building-permits], all building *Work Types* were included in the analysis, which may not accurately represent construction activity in each ward. Many recorded work types, such as renovations, demolitions, and minor repairs (e.g., HVAC installations or pipework), are not indicative of significant construction projects. Future research could refine the analysis by focusing on specific types of construction projects to better reflect actual development activity.

Additionally, as noted in [@sec-appendix], there are 45 unique statuses assigned to building permits, but this study analyzed only a subset of these statuses. This narrow focus may overlook important stages in the construction process. Expanding the analysis to include more permit statuses could offer a more comprehensive understanding of construction activity.

Another limitation is the lack of detailed data on specific projects within each budget category, which restricts insights into why certain wards receive more or less funding. Access to granular project-level data could provide deeper context and explanations for funding disparities.

Moreover, the analysis assumes linear relationships between population density, average household income, and budget allocations, which may oversimplify the underlying dynamics. Developing more complex models that account for non-linear relationships and interactions could yield a richer understanding of factors influencing construction activity.

Finally, the analysis assumes independence among wards, an assumption that may not hold true given shared infrastructure and resource dependencies. Future studies could incorporate spatial or network-based models to address interdependencies between wards.
